Stability and accuracy of free surface time integration in viscous flows

### 2016-08-03 free-surface-paper.tex
Typo.
Attempt to clarify a few things.

Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
of the domain. In this case, there can be flow in and out of the domain, so the boundary must move in time.
A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
matches the boundary conditions which one is trying to model, but it typically requires a deformable 
domain with frequent remeshing to avoid ill-conditioned cells.

@@

We would like to analyze the time evolution of the normal modes of this system: each mode 
is the relaxation of topography with a characteristic relaxation time.  
We denote the normal modes by $\left[ \mathbf{u}_i, p^\prime_i, \zeta_i \right]^T$, with
relaxation times $\tau_i$, where the subscript corresponds to the $i$th normal mode.

@@

\subsection{Accuracy and asymptotics of the scheme}
It is important to note that even though we derived the nonstandard finite difference scheme assuming a decaying exponential
for $\mathbf{u}$, it is formally a first-order accurate scheme. As such, it is capable of capturing arbitrary motions 
of the free surface, while allowing larger timesteps than forward Euler schemes.

@@

\begin{equation}
\begin{aligned}
a_i(\Delta t) - a_i^{1} &= a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
                        &= a_i{(0)} \left[ e^{-\Delta t/\tau_i} - 1 + \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right].
\end{aligned}
\end{equation}
Note that when $\tau_i = \tau^*$ the nonstandard finite difference scheme is exact.
The exponentials may be expanded to find
\ifdetail
\begin{equation}
a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right].
\end{equation}
\fi
\begin{equation}
a_i(\Delta t) - a_i^{1} = {a_i{(0)} } \left( \frac{\Delta t}{\tau_i} \right)^2 \left( 1 - \frac{\tau_i}{\tau^*} \right) + O(\Delta t^3).
\end{equation}
The local truncation error is quadratic in $\Delta t$, similar to the forward Euler, and summing this error over 
many timesteps results in a factor of $\Delta t^{-1}$, demonstrating the first-order global accuracy~\citep[e.g.][]{leveque2007finite}.
The choice of $\tau^*$ controls the size of the coefficient on the truncation error for the scheme.
The error for a given mode becomes considerably smaller when $\tau^*$ is close to its natural relaxation times.

@@

@@

\section{The relaxation spectrum is positive definite}
\label{app:positive_definite}
On physical grounds, we expect the relaxation spectrum described in Section~\ref{sec:eigenvalue}
to have only positive eigenvalues $\tau_i$.
An imaginary eigenvalue would have an oscillitory component,
which we do not expect in creeping flows (oscillation being usually due to inertia). 
A negative eigenvalue would grow exponentially, which would violate conservation of energy.
Here we demonstrate the positivity of the relaxation time spectrum.
We could proceed with either the continuous or the discretized version of the governing
equations, but choose the continuous form here.
We begin with the homogeneous Stokes equation (Equation~\eqref{eq:stokes}):
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
\label{eq:homogeneous_stokes}
\end{equation}
We can transform this into the mechanical energy equation by multiplying it by $\mathbf{u}$
and integrating over the domain:
\begin{equation}
\int_{\Omega(t)} \mathbf{u} \cdot \nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \int_{\Omega(t)} \mathbf{u}\cdot \nabla p = 0.
\end{equation}
Integrating by parts and applying boundary conditions yields
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Omega(t)}p \nabla \cdot \mathbf{u} - \int_{\Gamma_F(t)} \mathbf{u} \cdot \mathbf{T} \cdot \mathbf{n} = 0.
\end{equation}
Where the integrals correspond to the viscous dissipation, the work done by pressure, and the work done by boundary stresses, respectively.
Since we are using the incompressible approximation, the second integral is zero.
On a true free surface the boundary stresses are zero, but we make the same approximation
used in Equation~\eqref{eq:hydrostatic} and map the stress onto the reference surface, finding
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Gamma_F(t)} \rho_0 g \zeta \mathbf{u} \cdot \mathbf{n} = 0.
\end{equation}
We can now eliminate $\zeta$ from the equation in the same way that was used in Equation~\eqref{eq:weak_eigen},
thereby introducing the relaxation time $\tau_i$:
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2 = 0.
\end{equation}
This equation may be solved for $i\tau_i$:
\begin{equation}
\tau_i = \frac{\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})}{ \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2}.
\end{equation}
The top and bottom of the right hand side are both strictly positive numbers so the $\tau_i$ must also be positive,
demonstrating the positive definiteness of the system. A similar analysis can be performed for the discrete system.
There is the possibility for velocities which do not move the free surface at all, in which case the bottom
integral goes to zero, and the relaxation time becomes infinite. One example of this is the case of shear flow with
a free upper surface and periodic horizontal boundary conditions. Since we are usually interested in the smallest
relaxation times, these infinite eigenvalues are not usually a problem.

>>>>

Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
of the domain. In this case, there can be flow in and out of the domain, so the boundary must move in time.
A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
matches the boundary conditions which one is trying to model, but it typically requires a deformable 
domain with frequent remeshing to avoid ill-conditioned cells (i.e., cells that are inverted or have large aspect ratios).

@@

We would like to analyze the time evolution of the normal modes of this system: each mode 
corresponds to the relaxation of topography with a characteristic wavelength and relaxation time.  
These times are determined by the model geometry, gravity, viscosity structure, and density structure.
We denote the normal modes by $\left[ \mathbf{u}_i, p^\prime_i, \zeta_i \right]^T$, with
relaxation times $\tau_i$, where the subscript corresponds to the $i$th normal mode.

@@

\subsection{Accuracy and asymptotics of the scheme}
It is important to note that even though we derived the nonstandard finite difference scheme assuming a decaying exponential
for $\mathbf{u}$, it is formally a first-order accurate scheme. As such, it is capable of capturing arbitrary motions 
of the free surface, while allowing larger timesteps than the forward Euler scheme.

@@

\begin{equation}
\begin{aligned}
a_i(\Delta t) - a_i^{1} &= a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
                        &= a_i{(0)} \left[ e^{-\Delta t/\tau_i} - 1 + \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right].
\end{aligned}
\end{equation}
Note that when $\tau_i = \tau^*$ the nonstandard finite difference scheme is exact.
The exponentials may be expanded to find
\ifdetail
\begin{equation}
a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right].
\end{equation}
\fi
\begin{equation}
a_i(\Delta t) - a_i^{1} = {a_i{(0)} } \left( \frac{\Delta t}{\tau_i} \right)^2 \left( 1 - \frac{\tau_i}{\tau^*} \right) + O(\Delta t^3).
\end{equation}
The local truncation error is quadratic in $\Delta t$, similar to the forward Euler schem.
Summing this error over many timesteps results in a factor of $\Delta t^{-1}$, which results in a global truncation error that is linear in $\Delta t$ \citep[e.g.][]{leveque2007finite}, demonstrating the first-order accuracy.
The choice of $\tau^*$ controls the size of the coefficient on the truncation error for the scheme.
The error for a given mode becomes considerably smaller when $\tau^*$ is close to its natural relaxation times.

@@

The improved understanding of free surface boundary conditions in geodynamic models
will allow for modelers to make more informed choices about timestepping and free
surface stabilization techniques for use in simulations related to dynamic topography,
active tectonics, gravity anomalies, and geomorphologic processes.

@@

\section{The relaxation spectrum is positive definite}
\label{app:positive_definite}
On physical grounds, we expect the relaxation spectrum described in Section~\ref{sec:eigenvalue}
to have only positive eigenvalues $\tau_i$.
An imaginary eigenvalue would have an oscillitory component,
which we do not expect in creeping flows (oscillation being usually due to inertia). 
A negative eigenvalue would grow exponentially, which would violate conservation of energy.
Here we demonstrate the positivity of the relaxation time spectrum.
We could proceed with either the continuous or the discretized version of the governing
equations, but choose the continuous form here.
We begin with the homogeneous Stokes equation (Equation~\eqref{eq:stokes}):
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}.
\label{eq:homogeneous_stokes}
\end{equation}
We can transform this into the mechanical energy equation by multiplying it by $\mathbf{u}$
and integrating over the domain:
\begin{equation}
\int_{\Omega(t)} \mathbf{u} \cdot \nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \int_{\Omega(t)} \mathbf{u}\cdot \nabla p = 0.
\end{equation}
Integrating by parts and applying boundary conditions yields
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Omega(t)}p \nabla \cdot \mathbf{u} - \int_{\Gamma_F(t)} \mathbf{u} \cdot \mathbf{T} \cdot \mathbf{n} = 0.
\end{equation}
Where the integrals correspond to the viscous dissipation, the work done by pressure, and the work done by boundary stresses, respectively.
Since we are using the incompressible approximation, the second integral is zero.
On a true free surface the boundary stresses are zero, but we make the same approximation
used in Equation~\eqref{eq:hydrostatic} and map the stress onto the reference surface, finding
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \int_{\Gamma_F(t)} \rho_0 g \zeta \mathbf{u} \cdot \mathbf{n} = 0.
\end{equation}
We can now eliminate $\zeta$ from the equation in the same way that was used in Equation~\eqref{eq:weak_eigen},
thereby introducing the relaxation time $\tau_i$:
\begin{equation}
-\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})  + \tau_i \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2 = 0.
\end{equation}
This equation may be solved for $\tau_i$:
\begin{equation}
\tau_i = \frac{\int_{\Omega(t)} 2 \eta \varepsilon(\mathbf{u}) : \varepsilon(\mathbf{u})}{ \int_{\Gamma_F(t)} \rho_0 g (\mathbf{u} \cdot \mathbf{n})^2}.
\end{equation}
The top and bottom of the right hand side are both strictly positive numbers so the $\tau_i$ must also be positive,
demonstrating the positive definiteness of the system. A similar analysis can be performed for the discrete system.
There is the possibility for velocities which do not move the free surface at all, in which case the bottom
integral goes to zero, and the relaxation time becomes infinite. One example of this is the case of shear flow with
a free upper surface and periodic horizontal boundary conditions. Since we are usually interested in the smallest
relaxation times, these infinite eigenvalues are not usually a problem.
